---
title: "Untitled"
output: html_document
date: "2024-03-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Import packages
```{r, echo=TRUE}
library(tidyverse)
```
## Import data

```{r, echo=TRUE}
url_in <- "https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",
                 "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in, file_names)
urls
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[2])
```
## Tidy and analyze data

### Change date column to date type for both datasets cases and deaths
```{r, echo=TRUE}
US_cases <- US_cases %>%
pivot_longer(cols = -(UID:Combined_Key),
names_to = "date",
values_to = "cases") %>%
mutate(date = mdy(date))
```

```{r, echo=TRUE}
US_deaths <- US_deaths %>%
pivot_longer(cols = -(UID:Population),
names_to = "date",
values_to = "deaths") %>%
mutate(date = mdy(date))
```

### Filtering datasets to show Oregon data

```{r, echo=TRUE}
OR_cases <- US_cases %>%
  filter(Province_State == "Oregon") %>%
  group_by(Admin2, date) %>%
  summarize(total_cases = sum(cases))
```

```{r, echo=TRUE}
OR_deaths <- US_deaths %>%
filter(Province_State == "Oregon") %>%
group_by(Admin2, date) %>%
summarize(total_deaths = sum(deaths))
```

### Joining both datasets and filtering cases to more than a 0
```{r, echo=TRUE}
OR_totals <- left_join(OR_cases, OR_deaths, by = c("Admin2", "date"))
```

```{r, echo=TRUE}
OR_totals <- OR_totals %>%
filter(total_cases > 0)
```
### Calculating how much deaths happened per cases

```{r, echo=TRUE}
OR_deaths_per_cases <- OR_totals %>%
mutate(deaths_per_cases = total_deaths / total_cases)
```
## Visualization 1 - Deaths per Cases Over Time

```{r, echo=TRUE}
ggplot(OR_deaths_per_cases, aes(x = date, y = deaths_per_cases)) +
geom_line(color = "red") +
labs(x = "Date", y = "Deaths per Cases") +
ggtitle("Deaths per Cases Over Time")
```
### The visualiazation shows that overall deaths per cases have been decreasing since 2021 but there was a surge between 2022 and 2023.

#Visualization 2 - Top 5 Counite with Deaths per Cases

```{r, echo=TRUE}
OR_top_counties <- OR_deaths_per_cases %>%
group_by(Admin2) %>%
summarise(total_deaths_per_cases = sum(deaths_per_cases)) %>% arrange(desc(total_deaths_per_cases)) %>%
top_n(5)
```
```{r, echo=TRUE}
ggplot(OR_top_counties, aes(x = Admin2, y = total_deaths_per_cases)) +
geom_bar(stat = "identity", fill = "blue") +
labs(x = "County", y = "Most Deaths per Cases", title = "Top Counties by Most Deaths per Cases")
```
### This visualization indicates that Oregon counties Yamhill, Josephine, Wallowa, Linn, and Polk has the most deaths per cases in the state.

#Model

```{r, echo=TRUE}
model <- lm(deaths_per_cases ~ total_deaths + total_cases, data = OR_deaths_per_cases)
```

```{r, echo=TRUE}
OR_deaths_per_cases_predictions <- OR_deaths_per_cases %>%
ungroup() %>%
mutate(pred = predict(model))
```

```{r, echo=TRUE}
ggplot(OR_deaths_per_cases_predictions, aes(x = date)) +
geom_point(aes(y = deaths_per_cases), color = "blue", alpha = 0.5) +
geom_point(aes(y = pred), color = "red", alpha = 0.5) +
labs(x = "Date", y = "Deaths per Cases", title = "Actual vs.Predicted Deaths per Cases Over Time") +
theme_minimal()
```
### Overall the model predicts a slow but steady decrease of deaths per cases


## Bias
### As a resident or Oregon I have a personal investment in the COVID dynamic, particularly in reagrd to the rise of the anti-vaxing movements in certain areas of the state. So it was important to me to not look for a confirmation of my bias in this project.