---
title: "NYPD Incidents Historic Data Assessment"
date: "2024-03-05"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import Library

```{r, echo=TRUE}
library(tidyverse)
```

## Import Data

```{r, echo=TRUE}
NYPD_shooting <- read_csv("NYPD_Shooting_Incident_Data__Historic_.csv")
```
## Tidy and analyze data

### Changing the occured date column type to date

```{r, echo=TRUE}
NYPD_shooting <- NYPD_shooting %>%
mutate(OCCUR_DATE = mdy(OCCUR_DATE))
```

### Changing the borough column type to factor

```{r, echo=TRUE}
NYPD_shooting <- NYPD_shooting %>%
mutate(BORO = factor(BORO))
```

### Summarizing the total incidents per borough

```{r, echo=TRUE}
incidents_per_boro <- NYPD_shooting %>%
group_by(BORO) %>%
summarize(total_incidents = n())
```

### Finding out the 5 boroughs with the most shooting incidents

```{r, echo=TRUE}
top_5_boros <- incidents_per_boro %>%
arrange(desc(total_incidents)) %>%
slice(1:5)
```

### Summarizing the total incidents per date

```{r, echo=TRUE}
incidents_per_date <- NYPD_shooting %>%
group_by(OCCUR_DATE) %>%
summarize(total_incidents = n())
```

## Visualisation 1 - top 5 Boroughs with the most shooting incidents

```{r, echo=TRUE}
ggplot(top_5_boros, aes(x = BORO, y = total_incidents)) +
geom_bar(stat = "identity", fill = "skyblue") +
labs(title = "Top 5 Boroughs with the Most Incidents", x = "Borough", y = "Total Incidents") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Based on the analysis the top boroughs in New York area with NYPD shootings incidents are: Brooklyn, Bronx, Queens, Manhattan, Staten Island.

## Visualization 2 - number of incidents over time
```{r, echo=TRUE}
ggplot(incidents_per_date, aes(x = OCCUR_DATE, y = total_incidents)) +
geom_line(color = "blue") +
labs(title = "Number of Incidents Over Time", x = "Date", y = "Total Incidents")
```

### This visualization shows no obvious pattern, and the number of shooting incidents seem to go up and down but what is interesting is the surge in incidents after 2020.

## Model

```{r, echo=TRUE}
model <- lm(total_incidents ~ OCCUR_DATE, data = incidents_per_date)
```

```{r, echo=TRUE}
NYPD_shooting_predictions <- incidents_per_date %>%
ungroup() %>%
mutate(pred = predict(model))
```
```{r, echo=TRUE}
ggplot(NYPD_shooting_predictions, aes(x = OCCUR_DATE)) +
geom_point(aes(y = total_incidents), color = "blue", alpha = 0.5) +
geom_point(aes(y = pred), color = "red", alpha = 0.5) +
labs(x = "Date", y = "Deaths per Cases", title = "Actual vs.Predicted NYPD Shooting Incidents") + theme_minimal()
```
### The model shows a steady although not rapid decrease in the shooting incidents

## Bias

### My personal bias is influenced by the news information on the amount of crime in New York City, expecially it’s certain areas. WHen analyzing this data I tried not to look for a confirmaition of my bias but rather at what the data presents.